Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-14 analysis/paper/paper.qmd
working on figures

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Cores",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total",
           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
           )) %>% 
         filter(category != "Total"), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "#b2abd2",
    "Cores" = "gray"
  )) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = c(0.87, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black"),
    legend.box.margin = margin(0, 10, 0, 10)
  )

@@

```{r umap_by_category, include=FALSE}
category_colors <- c(
  "idealmodel" = "black",
  "spheroid" = "#e66101",
  "subspheroid" = "#fdb863",
  "multifacial" = "#b2abd2",
  "polyhedron" = "#5e3c99"
)

ggplot(umap_data, aes(x = x, y = y)) +
  geom_point(aes(color = category_lower, size = point_size), alpha = 0.9, shape = 16) +
  #geom_text_repel(
    #data = label_data,
    #aes(label = filename),
    #size = 3,
    #max.overlaps = 100,
    #show.legend = FALSE
  #) +
  scale_color_manual(values = category_colors, name = "Category") +
  scale_size_identity() +
  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded)+ #确保范围一致
  theme_minimal(base_size = 14) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  ) +
  theme(
    legend.position = "non",
    # plot.title = element_text(face = "bold")
  )
```

```{r umap_by_cluster, include=FALSE}
cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")

ggplot(umap_data, aes(x = x, y = y)) +
  geom_point(
    data = filter(umap_data, is_idealmodel == "no"),
    aes(color = cluster),
    alpha = 0.8,
    size = 2.5
  ) +
  geom_point(
    data = filter(umap_data, is_idealmodel == "yes"),
    aes(color = cluster),
    size = 3.5,
    show.legend = FALSE
  ) +
  stat_ellipse(
    aes(color = cluster),
    type = "norm",
    level = 0.95,
    linetype = "dashed",
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = cluster_colors,
    name = "Cluster"
  ) +
  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded) +  # 保证坐标一致
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    #caption = "Dashed ellipses show 95% confidence intervals"
  )
```

@@

```{r}
#| label: fig-UMAP-combined
#| fig-cap: "A. UMAP plot showing distinct clusters of shapes derived from SPHARM. B. UMAP plot of SPHARM-derived shape features colored by types"
#| echo: false

>>>>

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Cores",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total",
           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
           )) %>% 
         filter(category != "Total"), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "#b2abd2",
    "Cores" = "gray"
  )) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = c(0.88, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black"),
    legend.box.margin = margin(0, 10, 0, 10)
  )

@@

```{r}
#| echo: false
#| label: fig-umap-by-category
#| include: false
#| 
category_colors <- c(
  "ideal model" = "black",
  "spheroid" = "#e66101",
  "subspheroid" = "#fdb863",
  "multifacial" = "#b2abd2",
  "polyhedron" = "#5e3c99"
)

ggplot(umap_data %>% 
         mutate(category =
                if_else(category == "idealmodel",
                       "ideal model",
                       category)) %>% 
         mutate(category_lower =
                if_else(category_lower == "idealmodel",
                       "ideal model",
                       category_lower)), 
       aes(x = x, y = y)) +
  geom_point(aes(color = category_lower, 
                 size = point_size), 
             alpha = 0.9, 
             shape = 16) +
  scale_color_manual(values = category_colors, 
                     name = "Category") +
  scale_size_identity() +
  coord_cartesian(xlim = x_limits_expanded, 
                  ylim = y_limits_expanded)+ #确保范围一致
  theme_minimal(base_size = 14) +
  theme(legend.position = c(.2,.75),
         legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
         legend.box.background = element_rect(color = "black"),
         legend.box.margin = margin(0, 0, 0, 0)) +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  ) 
```

```{r}
#| echo: false
#| label: fig-umap-by-cluster
#| include: false

cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")
library(ggforce)

ggplot(umap_data %>%  
         mutate(category_lower =
                if_else(category_lower == "idealmodel",
                       "ideal model",
                       category_lower)), 
       aes(x = x, y = y)) +
  geom_point(
    data = filter(umap_data, is_idealmodel == "no"),
    aes(color = cluster),
    alpha = 0.8,
    size = 2.5
  ) +
  geom_point(
    data = filter(umap_data, 
                  is_idealmodel == "yes"),
    aes(color = cluster),
    size = 3.5,
    show.legend = FALSE
  ) +
  geom_mark_ellipse(aes(color = cluster, 
                        label = category)) +
  #stat_ellipse(
  #  aes(color = cluster),
  #  type = "norm",
  #  level = 0.95,
  #  linetype = "dashed",
  #  show.legend = FALSE
  #) +
  scale_color_manual(
    values = cluster_colors,
    name = "Cluster"
  ) +
  coord_cartesian(xlim = x_limits_expanded, 
                  ylim = y_limits_expanded) +  # 保证坐标一致
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  )
```

@@

```{r}

#| label: fig-UMAP-combined
#| fig-cap: "A. UMAP plot showing distinct clusters of shapes derived from SPHARM. B. UMAP plot of SPHARM-derived shape features colored by types"
#| echo: false
